---
title: "The Changing Dynamics of Central American Migration"
author: "Mario Moreno"
date: "2/3/2019"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---

```{r setup, include=FALSE}
chooseCRANmirror(graphics=FALSE, ind=1)
knitr::opts_chunk$set(echo = TRUE)
```
```{r packsandlibs, include=FALSE, warning=FALSE}
install.packages('rmarkdown')
install.packages('knitr')
install.packages('ggplot2')
install.packages('tidyverse')
install.packages('gghighlight')
install.packages('ggalluvial')


library(ggalluvial)
library(ggplot2)
library(ggrepel)
library(gghighlight)
library(gtools)
library(tidyverse)
library(dplyr)
```
```{r dataloads, include=FALSE}
# READING IN FOR BAR, LINE, AND BUBBLE GRAPHS
elsLAPOP <- read.csv('Data/elsLAPOP.csv', encoding='latin1')
guateLAPOP <- read.csv('Data/guateLAPOP.csv', encoding='latin1')
hondLAPOP <- read.csv('Data/hondLAPOP.csv', encoding = 'latin1')
sec <- read.csv('Data/security.csv')

elsLAPOP$pais[elsLAPOP$pais==3] <- "El Salvador"
guateLAPOP$pais[guateLAPOP$pais==2] <- "Guatemala"
hondLAPOP$pais[hondLAPOP$pais==4] <- "Honduras"

cols_to_keep <- intersect(colnames(elsLAPOP), colnames(guateLAPOP))
g <- guateLAPOP[, cols_to_keep, drop=FALSE]
e <- elsLAPOP[, cols_to_keep, drop=FALSE]
bind <- smartbind(g, e)
cols_to_keep <- intersect(colnames(bind), colnames(hondLAPOP))
h <- hondLAPOP[, cols_to_keep, drop=FALSE]
b <- bind[, cols_to_keep, drop=FALSE]
LAPOP <- smartbind(h, b)

rm(cols_to_keep, g, e, h, b, bind)

## READING FOR ALLUVIAL GRAPH
hmex <- read.csv('Data/honddev2017mex.csv')
hmex$pais <- 'Honduras'
hmex$deported <- 'Mexico'

husa <- read.csv('Data/honddev2017usa.csv')
husa$pais <- 'Honduras'
husa$deported <- 'USA'

cols_to_keep_mex <- c('pais', 'deported', 'p15', 'p22_1e', 'p22_1l', 'p22_2e', 'p22_2l', 'p24l', 'p34l')
cols_to_keep_usa <- c('pais', 'deported', 'p14_1', 'p19_1e', 'p19_1l', 'p19_2e', 'p19_2l', 'p20l', 'p26')

husa1 <- husa[cols_to_keep_usa]
hmex1 <- hmex[cols_to_keep_mex]

colnames(husa1)[colnames(husa1)=="p14_1"] <- "p15"
colnames(husa1)[colnames(husa1)=="p19_1e"] <- "p22_1e"
colnames(husa1)[colnames(husa1)=="p19_1l"] <- "p22_1l"
colnames(husa1)[colnames(husa1)=="p19_2e"] <- "p22_2e"
colnames(husa1)[colnames(husa1)=="p19_2l"] <- "p22_2l"
colnames(husa1)[colnames(husa1)=="p20l"] <- "p24l"

alluvial <- dplyr::bind_rows(husa1, hmex1)
alluvial[is.na(alluvial)] <- 0
rm(hmex, husa, hmex1, husa1, cols_to_keep_usa, cols_to_keep_mex)

## READING IN FOR SLOPE GRAPH
path <- "Data/"
files <- list.files(path=path, pattern="*usa.csv")
for(file in files)
{
  perpos <- which(strsplit(file, "")[[1]]==".")
  assign(
    gsub(" ","",substr(file, 1, perpos-1)), 
    read.csv(paste(path,file,sep="")))
}

# Add country identifiers
elsdev2014usa$pais <- 'Price paid by El Salvadorean migrants'
elsdev2015usa$pais <- 'Price paid by El Salvadorean migrants'
elsdev2016usa$pais <- 'Price paid by El Salvadorean migrants'
elsdev2017usa$pais <- 'Price paid by El Salvadorean migrants'

honddev2014usa$pais <- 'Price paid by Honduran migrants'
honddev2015usa$pais <- 'Price paid by Honduran migrants'
honddev2016usa$pais <- 'Price paid by Honduran migrants'
honddev2017usa$pais <- 'Price paid by Honduran migrants'

guatedev2014usa$pais <- 'Price paid by Guatemalan migrants'
guatedev2015usa$pais <- 'Price paid by Guatemalan migrants'
guatedev2016usa$pais <- 'Price paid by Guatemalan migrants'
guatedev2017usa$pais <- 'Price paid by Guatemalan migrants'

# subset dataframes
e2014 <- elsdev2014usa[c('year', 'pais', 'p17_1c', 'p17_1u')]
e2015 <- elsdev2015usa[c('year', 'pais', 'p17_1c', 'p17_1u')]
e2016 <- elsdev2016usa[c('year', 'pais', 'p17_1c', 'p17_1u')]
e2017 <- elsdev2017usa[c('year', 'pais', 'p17_1c', 'p17_1u')]

h2014 <- honddev2014usa[c('year', 'pais', 'p17_1c', 'p17_1u')]
h2015 <- honddev2015usa[c('year', 'pais', 'p17_1c', 'p17_1u')]
h2016 <- honddev2016usa[c('year', 'pais', 'p17_1c', 'p17_1u')]
h2017 <- honddev2017usa[c('year', 'pais', 'p17_1c', 'p17_1u')]

g2014 <- guatedev2014usa[c('year', 'pais', 'p17_1c', 'p17_1u')]
g2015 <- guatedev2015usa[c('year', 'pais', 'p17_1c', 'p17_1u')]
g2016 <- guatedev2016usa[c('year', 'pais', 'p17_1c', 'p17_1u')]
g2017 <- guatedev2017usa[c('year', 'pais', 'p17_1c', 'p17_1u')]

rm(elsdev2014usa, elsdev2015usa, elsdev2016usa, elsdev2017usa, honddev2014usa, honddev2015usa, honddev2016usa, honddev2017usa, guatedev2014usa, guatedev2015usa, guatedev2016usa, guatedev2017usa)

# bind them 
slopegraph <- do.call("rbind", list(e2014, e2015, e2016, e2017, h2014, h2015, h2016, h2017, g2014, g2015, g2016, g2017))
rm(e2014, e2015, e2016, e2017, h2014, h2015, h2016, h2017, g2014, g2015, g2016, g2017)

# READ IN NIGHTINGALE

hmex <- read.csv('Data/honddev2017mex.csv')
hmex$pais <- 'Honduras'
hmex$deported <- 'Mexico'

husa <- read.csv('Data/honddev2017usa.csv')
husa$pais <- 'Honduras'
husa$deported <- 'USA'

gmex <- read.csv('Data/guatedev2017mex.csv')
gmex$pais <- 'Guatemala'
gmex$deported <- 'Mexico'

gusa <- read.csv('Data/guatedev2017usa.csv')
gusa$pais <- 'Guatemala'
gusa$deported <- 'USA'

emex <- read.csv('Data/elsdev2017mex.csv')
emex$pais <- 'El Salvador'
emex$deported <- 'Mexico'

eusa <- read.csv('Data/elsdev2017usa.csv')
eusa$pais <- 'El Salvador'
eusa$deported <- 'USA'

usa1 <- smartbind(eusa, gusa)
usa <- smartbind(usa1, husa)
rm(usa1)

mex1 <- smartbind(emex, gmex)
mex <- smartbind(mex1, hmex)
rm(mex1)

rm(husa, hmex, gmex, gusa, emex, eusa, usa1, mex1)
```
```{r themes, include=FALSE}
mario_theme <- theme(text = element_text(family='sans'),
                     plot.title = element_text(color='#7a0177', size=14, face='bold'),
                     plot.subtitle = element_text(color='black', size=12, face='italic'), 
                     axis.title.x = element_text(color='black', size=10),
                     axis.title.y = element_text(color='black', size=10), 
                     plot.caption = element_text(color='grey', size=8),
                     legend.title = element_text(face='bold', size=8),
                     legend.position = 'bottom', legend.direction = 'horizontal', legend.box = 'horizontal', 
                     legend.key.size = unit(0.5, 'cm'))

mario_scale_cont <- scale_fill_gradient(low='#feebe2', high='#7a0177')
mario_manual <- scale_color_manual(values=c('#DDA0DD','#FF00FF', '#4B0082'))
```


In the midst of a raging national debate on migration and refugees, plenty has been written about the changing profile of undocumented migrants reaching the US-Mexico border. While traditionally, Mexican migrants have been the predominant demographic group seeking to enter the United States through the southern border, that dynamic has shifted in the last decade. Central American migrants originating in the Northern Triangle countries of El Salvador, Guatemala, and Honduras have become the predominant demographic of undocumented migrants seeking to enter the United States. 

```{r refugees, warning=FALSE}
# Calulcate the percent of people who've considered emigration over time

inter <- LAPOP %>% 
  select(year, pais, q14) %>%
  group_by(year, pais) %>%
  add_tally() %>%
  add_count(q14)

inter$perc <- (inter$nn / inter$n) * 100

to_graph <- inter %>%
  select(year, pais, perc, q14) %>%
  group_by(year, pais) %>%
  arrange(q14) %>%    
  slice(1)

# Plot
 
plot <- ggplot(to_graph, aes(x=as.factor(year), y=pais, fill=perc)) + geom_tile(aes(fill=perc)) + geom_text(aes(label=round(perc, 1), fontface='bold')) + mario_scale_cont 

labels <- labs(title = 'Since 2012, the percentage of Honduran residents \n planning to emigrate has increased nearly 30%',
               subtitle = 'Across the region, there are notable increases in plans to emigrate since 2012',
               caption = 'Source: LAPOP 2010-2016',
               x = 'Year',
               y = 'Country',
               fill = '% planning to emigrate')

update_theme <- theme(axis.title.x = element_blank(),
                      axis.title.y = element_blank(),
                      panel.grid = element_blank(), 
                      axis.ticks = element_blank(),
                      panel.background = element_blank())

plot + labels + mario_theme + update_theme

rm(inter, to_graph, plot)
```

As evidenced by the graph above, this surge in Central American migration is evident in the percentage of citizens of Central American countries who are planning to emigrate in the next three years. Notably, in all three countries, the percentage of citizens planning to emigrate has considerably increased since 2012 -- with the percentage of Hondurans planning to make the trip out of the country rising from 11% in 2012 to 40% in 2016. 

Media outlets and researchers often cite a range of push and pull factors driving this emigration. Generally speaking, violence and lack of economic opportunity are cited as two prominent push factors; while family reunification is often cited as a pull factor. 

```{r drivers, warning=FALSE}
## This graph will show perceptions in violence and economic opportunity
## relative to answers on willingness to emigrate in the next 3 years for 2017

hond2017 <- subset(hondLAPOP, year==2016)

# Calculate the percent of people in each province who said they felt
# "Very unsafe" in the community

unsafe <- hond2017 %>%
  select(year, prov, aoj11) %>%
  group_by(year, prov) %>%
  add_tally() %>%
  add_count(aoj11)

# Calculate the percent of people in each province who said their economic
# situation has worsnened in the last 12 months

econ <- hond2017 %>%
  select(year, prov, idio2) %>%
  group_by(year, prov) %>%
  add_tally() %>%
  add_count(idio2)

# Calculate the percent of people in each province who want to emigrate
# in the next 3 years

emigrate <- hond2017 %>%
  select(year, prov, q14) %>%
  group_by(year, prov) %>%
  add_tally() %>%
  add_count(q14)

# Group by and replace those provinces that don't have those values with zero
# percent

emigrate$perc_emigrate <- (emigrate$nn / emigrate$n) * 100
unsafe$perc_aoj11 <- (unsafe$nn / unsafe$n) * 100
econ$perc_idio2 <- (econ$nn / econ$n) * 100

emigrate.na <- na.omit(emigrate)
unsafe.na <- na.omit(unsafe)
econ.na <- na.omit(econ)

to_plot_emigrate <- emigrate.na %>%
  select(year, prov, perc_emigrate, q14) %>%
  group_by(year, prov) %>%
  arrange(q14) %>%
  slice(1)

to_plot_econ <- econ.na %>%
  select(year, prov, perc_idio2, idio2) %>%
  group_by(year, prov) %>%
  arrange(desc(idio2)) %>%    
  slice(1)

to_plot_unsafe <- unsafe.na %>%
  select(year, prov, perc_aoj11, aoj11) %>%
  group_by(year, prov) %>%
  arrange(desc(aoj11)) %>%    
  slice(1)

to_plot_final <- merge(to_plot_econ, to_plot_unsafe, by = c("year", "prov"))
to_plot_final <- merge(to_plot_final, to_plot_emigrate, by = c("year", "prov"))
to_plot_f <- to_plot_final %>% mutate(perc_aoj11 = ifelse(aoj11 != 4, 0, perc_aoj11))
to_plot_ff <- to_plot_f %>% mutate(perc_idio2 = ifelse(idio2 != 3, 0, perc_idio2))

rm(to_plot_final, to_plot_f, to_plot_econ, to_plot_unsafe, emigrate, unsafe, econ, unsafe.na, econ.na, emigrate.na, hond2017, to_plot_emigrate)                                   

# Plot


plot <- ggplot(to_plot_ff, 
               aes(y = perc_aoj11, x = perc_idio2, color = perc_emigrate)) + geom_point() +          
  geom_text_repel(aes(label=prov)) + scale_color_distiller(name = '% planning to emigrate', palette='RdPu', trans='reverse') #+ 
  #scale_y_continuous(limits = c(0, 60)) + expand_limits(x=0)

labels <- labs(title = "In Honduran provinces with largest percentage of people planning \n to emigrate, economic concerns prevail",
            subtitle = "Economics might be playing a larger push role than violence",
            caption = "Source: LAPOP Honduras data 2017", 
            x = "% who feel economic prospects have worsened", y = "% who feel very unsafe from crime",
            fill='Province') 


plot + labels + mario_theme 

rm(to_plot_ff, plot)
```

However, as the graph above shows, it appears that in Honduras -- the country that's seen the largest increases in its citizens planning to emigrate -- a lack of economic opportunity is driving emigration. In the Honduran states where more than 40% of survey respondents plan to emigrate in the next three years, it's clear that more respondents feel their economic prospects have worsened. In the state of Colón for instance, more than 50% of the population wants to emigrate, more than 50% feel that their economic prospects have worsened, and only 20% feel that their neighborhood is very unsafe from crime. 

Even as it appears that Honduran migrants are fleeing poverty more than violence, they face additional economic burdens as they try to make it to the United States. Often, in order to secure a safe passage from the Guatemala-Mexico border to the US-Mexico border, these migrants will secure the assistance of 'coyotes,' members of criminal organizations that guide them in their journey. 


```{r cost, warning = FALSE}
tograph <- slopegraph %>%
  group_by(year, pais) %>%
  filter(p17_1c >0, p17_1u == 3) %>%
  summarize(mean_spent = mean(p17_1c)) %>%
  mutate_if(is.numeric, round, 0)

rm(e2014, e2015, e2016, e2017, h2014, h2015, h2016, h2017, g2014, g2015, g2016, g2017)

# Slope graph

 
plot <- ggplot(data = tograph, aes(x = year, y = mean_spent, group = pais)) +
  geom_line(aes(color = pais, alpha = 1), size = 1) +   mario_manual +
  geom_text_repel(data = tograph %>% filter(year == 2017), 
                  aes(label = pais) , 
                  hjust = "right", 
                  fontface = "bold", 
                  size = 3.5, 
                  nudge_x = -0.25, 
                  direction = "y") + 
  geom_label(aes(label = paste0("$", mean_spent)), 
             size = 3, 
             label.padding = unit(0.05, "lines"), 
             label.size = 0.0) 

updated_theme <- 
  # move the x axis labels up top
  #scale_x_discrete(position = "top") +
  # Format tweaks
  # Remove the legend
  theme(legend.position = "none") +
  # Remove the panel border
  theme(panel.border     = element_blank()) +
  # Remove just about everything from the y axis
  theme(axis.title.y     = element_blank()) +
  theme(axis.text.y      = element_blank()) +
  theme(panel.grid.major.y = element_blank()) +
  theme(panel.grid.minor.y = element_blank()) +
  # Remove a few things from the x axis and increase font size
  theme(axis.title.x     = element_blank()) +
  theme(panel.grid.major.x = element_blank()) +
  theme(axis.text.x.top      = element_text(size=12)) +
  # Remove x & y tick marks
  theme(axis.ticks       = element_blank()) + theme(panel.background = element_blank())
  #  Labelling as desired
labels <- labs(
    title = "Honduran migrants have experienced marked increases in the price of \n hiring a coyote to guide them through Mexico",
    subtitle = "Guatemalan migrants have experienced similar increases, but Salvadorean migrants \n have paid less for a successful trip",
    caption = "Source: EMIF Sur (2014-2017)"
  )

plot + labels + mario_theme + updated_theme

rm(tograph)
```

As evidenced by the graph above, these coyotes are expensive -- and are getting more costly over time. For migrants originating from the three Northern Triangle countries who safely made it to the US-Mexico border but got deported from the United States, the price paid to a coyote has increased significantly. Nowhere is this increase more prominent than for the Honduran migrants who are fleeing dire economic circumstances. The cost of a coyote for those Honduran migrants who reached the US-Mexico border has increased from 3942 dollars in 2014 to 6386 dollars in 2017. 

Even with the cost associated with the trip, migrants face signifincant dangers in their journey through Mexico. As they traverse a country known for it's powerful criminal organizations, associated violence, and weak protections for migrants, these migrants are often robbed, injured, or more. 

```{r challenges, warning = FALSE}
## Get data into percentages of each event
usa_g <- usa %>% 
  select(sexo, pais, p18_1, p18_2, p18_3, p18_4, p18_5, p18_6, p18_7, p18_8, p18_9, p18_10, p18_11, p18_12, p18_13) %>%
  group_by(sexo, pais) %>%
  filter(p18_1 > 0, p18_2 > 0, p18_3 >0, p18_4 > 0, p18_5 > 0, p18_6 > 0, p18_7 > 0, p18_8 > 0, p18_9 > 0, p18_10 > 0, p18_11 > 0, p18_12 > 0, p18_13 > 0) %>%
  filter(p18_1 < 10, p18_2 < 10, p18_3 <10, p18_4 < 10, p18_5 < 10, p18_6 < 10, p18_7 < 10, p18_8 < 10, p18_9 < 10, p18_10 < 10, p18_11 < 10, p18_12 < 10, p18_13 < 10)
usa_g[usa_g==2] <- 0

# since it's 0 and 1, a mean will work
usa_perc <- usa_g %>%
  group_by(sexo, pais) %>%
  summarise_all(mean) %>%
  select(sexo, pais, p18_1, p18_2, p18_6, p18_8, p18_9) 

# Rename columns
colnames(usa_perc)[colnames(usa_perc)=="p18_1"] <- "Extreme Cold"
colnames(usa_perc)[colnames(usa_perc)=="p18_2"] <- "Lack of Food"
colnames(usa_perc)[colnames(usa_perc)=="p18_6"] <- "Lost"
colnames(usa_perc)[colnames(usa_perc)=="p18_8"] <- "Abandoned by Coyote"
colnames(usa_perc)[colnames(usa_perc)=="p18_9"] <- "Assault or Robbery"

# Melt
library(reshape2)
melt <- melt(usa_perc, id.vars=c('sexo', 'pais'))
melt$value <- melt$value * 100

# Nightingale graph

plot <- ggplot(melt, aes(x=variable, y=value, fill=pais)) + geom_bar(width = 1, stat='identity') + 
  coord_polar(start = 1, direction=1, clip = 'off') +
  scale_fill_brewer(palette='RdPu')+ xlab("") + ylab("Percent facing challenge")

labels <- labs(title='Honduran migrants are more likely to be robbed, \n while Salvadorean migrants face extreme cold',
subtitle='Hondurans pay more for coyotes, but are \n the only group likely to be abandoned by them or get lost')

plot + labels + mario_theme 

```

In the graph above, five of the most common self-reported challenges facing migrants who made it to the United States in 2017 are shown. Notably, Honduran migrants faced the largest share of assault or robberies of all, while Salvadorean migrants faced a disproportionate share of extreme cold and lack of food. Interestingly, despite paying more for coyotes to guide them through Mexico, Honduran migrants are the only group that reported getting lost or getting abandoned by their coyote on the trip.

In addition to these challenges, there is guarantee that Central American migrants will evade migration forces in Mexico. The likelihood of being deported while traversing Mexico has likely increased in the last three years. Mexico, working in partnership with the United States, has beefed up its efforts to deport Central American migrants since 2014. 


```{r routes, warning = FALSE}

drop1 <- alluvial %>% filter(p15 >= 0, p22_1l >= 0, p22_2l >= 0, p26 >= 0, p24l >= 0)
clean <- drop1 %>% filter(p15 < 2000000, p22_1l < 900000000, p22_2l < 900000000, p26 < 300000000, p24l < 9000000000)
rm(drop1)

alluvialnodeshond <- rename(count(clean, deported, p15, p22_1l, p22_2l, p24l, p26, p34l), Freq = n)
touse <- alluvialnodeshond %>% filter(Freq>2)
rm(clean, alluvialnodeshond)

## Make names for it

touse$p15[touse$p15 == 1217001] <- 'Tecun Uman'
touse$p15[touse$p15 == 1705099] <- 'Naranjo'
touse$p15[touse$p15 == 1705016] <- 'El Ceibo'
touse$p15[touse$p15 == 1705015] <- 'Bethel'
touse$p15[touse$p15 == 1312047] <- 'La Mesilla'

touse$p22_1l[touse$p22_1l == 301930001] <- 'Veracruz, VER'
touse$p22_1l[touse$p22_1l == 300390001] <- 'Coatzacoalcos, VER'
touse$p22_1l[touse$p22_1l == 270170001] <- 'Tenosique de Pino Suarez, TAB'
touse$p22_1l[touse$p22_1l == 270040001] <- 'Villahermosa, TAB'
touse$p22_1l[touse$p22_1l == 240280001] <- 'San Luis Potosi, SLP'
touse$p22_1l[touse$p22_1l == 190390001] <- 'Monterrey, NL'
touse$p22_1l[touse$p22_1l == 70650001] <- 'Palenque, CHIS'
touse$p22_1l[touse$p22_1l == 70270001] <- 'Chiapa de Corzo, CHIS'


touse$p24l[touse$p24l == 300030001] <- 'Acayucan, VER'
touse$p24l[touse$p24l == 280320001] <- 'Reynosa, TAMPS'
touse$p24l[touse$p24l == 280270001] <- 'N. Laredo, TAMPS'
touse$p24l[touse$p24l == 270020001] <- 'Cardenas, TAB'
touse$p24l[touse$p24l == 301930001] <- 'Veracruz, VER'
touse$p24l[touse$p24l == 300390001] <- 'Coatzacoalcos, VER'
touse$p24l[touse$p24l == 270170001] <- 'Tenosique de Pino Suarez, TAB'
touse$p24l[touse$p24l == 270040001] <- 'Villahermosa, TAB'
touse$p24l[touse$p24l == 70650001] <- 'Palenque, CHIS'

touse$p34l[touse$p34l == 0] <- 'Avoided Deportation'
touse$p34l[touse$p34l == 300030001] <- 'Acayucan, VER'
touse$p34l[touse$p34l == 270020001] <- 'Cardenas, TAB'
touse$p34l[touse$p34l == 301930001] <- 'Veracruz, VER'
touse$p34l[touse$p34l == 300390001] <- 'Coatzacoalcos, VER'
touse$p34l[touse$p34l == 270170001] <- 'Tenosique de Pino Suarez, TAB'
touse$p34l[touse$p34l == 70650001] <- 'Palenque, CHIS'
touse$p34l[touse$p34l == 270040001] <- 'Villahermosa, TAB'


touse$p26[touse$p26 == 0] <- 'Deported Prior'
touse$p26[touse$p26 == 280320001] <- 'Reynosa, TAMPS'
touse$p26[touse$p26 == 280270001] <- 'N. Laredo, TAMPS'

colnames(touse)[colnames(touse)=="p15"] <- "entry"


plot <- ggplot(touse,
       aes(y = Freq,
           axis1 = entry, axis2 = p22_1l, axis3 = p24l, axis4 = p26)) +   scale_fill_brewer(type='qual', palette = "Set1") +
  geom_flow(aes(fill = deported, alpha = entry, colour = deported), stat = "alluvium", aes.bind = TRUE, reverse=TRUE, width = 0.1) +
  geom_stratum(width = 1/10, reverse = TRUE) +
  geom_text(stat = "stratum", label.strata = TRUE, reverse = TRUE) +
  scale_x_continuous(breaks = 1:4, labels = c("Mexican Port of Entry", "Mexican City Visited", "Mexican City Where Spent Most Time" ,"US Port of Entry")) 

labels <- labs(
    title = "As they journey northward, Honduran migrants who get deported \n in Mexico get stuck in cities that border the Gulf of Mexico",
    subtitle = "Most migrants who make it to the US enter through Reynosa",
    caption = "Source: EMIF Sur (2014-2017)") 

updated_theme <- theme(panel.background = element_blank(),
                       axis.title.y = element_blank(),
                       axis.text.y = element_blank(),
                       axis.ticks = element_blank())

plot+ labels + mario_theme + updated_theme

rm(touse)
```

The graph above, which shows the paths that Honduran migrants take within Mexico, shows that there are some slight differences in the paths that deported migrants and non-deported migrants take. Namely, migrants entering Mexico through Bethel tend to all be deported, while migrants who make a stop in Chiapa de Corso tend to all make it to the US-Mexico border. Additionally, it seems that deported migrants tend to be caught in cities either near the Guatemala-Mexico border or on the Gulf of Mexico, while migrants that manage to make it safely through those cities often wind up making it to Reynosa, where they cross into the United States. 

In addition to deportation, Central American migrants working their way through Mexico face significant dangers. For that reason, academic experts and migrants rights advocates have often called for more investment in Northern Triangle countries to reduce violence, create more economic opportunity, and incentivize potential migrants to stay home. 


```{r usaid, echo=FALSE, warning = FALSE}
hond_sec <- subset(sec, country_name == 'Honduras')
hond_sec$dac_sector_name[is.na(hond_sec$dac_sector_name)] <- "Conflict, Peace, and Security"

to_plot_sec <- hond_sec %>%
  select(fiscal_year, dac_sector_name, constant_amount) %>%
  group_by(fiscal_year, dac_sector_name) %>%
  tally(constant_amount)

plot <- gghighlight_line(to_plot_sec, aes(fiscal_year, n, colour = dac_sector_name), predicate = max(n),
                 max_highlight = 3) + scale_y_continuous(name = "Amount (in 10,000,000)", labels = c(0, 2, 4, 6, 8)) + mario_manual

labels <- labs(title = "US Investments in Honduras have focused on civil society \n and government",
             subtitle = "There's notable increases since 2015, the year after the unaccompanied minors crisis",
             caption = "Source: USAID Foreign Aid Explorer and Security Assistance Monitor data, 2010-2017", 
             x = "Fiscal Year",
             fill='Province')  

plot + labels + mario_theme

rm(plot, to_plot_sec, hond_sec)
```

To that end, as the graph above shows, US aid to Honduras has emphasized investments in civil society and government since 2010. Notably, investments in these areas increased significantly in 2015, likely as a response to the unaccompanied minors crisis on the US southern border. Though, given the surge in Hondurans planning to emigrate in the next three years, it appears this aid has yet to make an impact in the country.  
